Inferring the gravitational potential of the Milky Way with a few precisely measured stars

### 2014-03-16 text/paper2/ms.tex
added choi reference
add figure references

% packages
\usepackage{amssymb,amsmath, amsbsy}
\usepackage{booktabs}

@@

To model the true, 6D, present-day position of the star, $\W$, we first transform to cartesian, Galactocentric coordinates where ($\bs{r},\bs{v}$) and ($\bs{r}_\sat,\bs{v}_\sat$) are the position and velocity of the star and progenitor. In cartesian coordinates, the present-day position of the star is taken as having been sampled at $t=\tub$ from an isotropic Gaussian centered on one of the instantaneous, effective Lagrange points in position, and an isotropic Gaussian centered on the origin in velocity, and evolved to present day in the parent potential, $\Phi$. We assume that once the star becomes ``unbound'' from the satellite the potential of the satellite can be ignored, and thus treat the star as a test particle.\footnote{This assumption breaks down for sufficiently large-mass progenitors, somewhere between $\sim10^8-10^9~$M$_\odot$.} The satellite mass enters the model through the tidal radius; a changing satellite mass will skew the positions of the effective Lagrange points and velocity scale. We allow for mass-loss by incorporating the initial satellite mass, $m$, and a constant mass-loss term, $\dot{m}$ into the model. Finally, we add two additional parameters: a constant, global offset, $\Loffset$, that scales the position of the effective Lagrange points relative to the classical tidal radius, and a binary parameter $\tailbit$, that is either $-1$ or $+1$ depending on whether the star is in the leading or trailing tail. The likelihood is then:
\begin{align}
	p(\W \given m, \dot{m}, \W_\sat, \tub, \Phi, \Loffset, \tailbit) &= p(\bs{r} \given \bs{r}_\sat, m, \dot{m}, \tub, \Phi, \Loffset, \tailbit)\,p(\bs{v} \given \bs{v}_\sat, m, \dot{m}, \tub, \Phi)\,\left\vert\J\right\vert_{t=\tub}\\
	p(\bs{r} \given \bs{r}_\sat, m, \dot{m}, \tub, \Phi, \Loffset, \tailbit) &= \mathcal{N}(\bs{r} \given \bs{r}_\sat - \Loffset\,\tailbit\,\rtide\,\hat{x}_1, \sigma^2_r)_{t=\tub}\\
	p(\bs{v} \given \bs{v}_\sat, m, \dot{m}, \tub, \Phi) &= \mathcal{N}(\bs{v} \given \bs{v}_\sat, \sigma^2_v)_{t=\tub}
\end{align}
where $\rtide$ is given by Eq.~\ref{eq:tidalradius}, and $\sigma_v$ by Eq.~\ref{eq:velscale}. $\left\vert\J\right\vert$ is the absolute value of the determinant of the Jacobian that defines the transformation from heliocentric, spherical to Galactocentric, cartesian coordinates. For a single star, in detail, the model works as follows:
\begin{enumerate}
	\item sample from the observed distributions in heliocentric coordinates (Eq.~\ref{eq:obsstar} and Eq.~\ref{eq:obsprog}) to get present-day positions for both the star and progenitor;
	\item transform to Galactocentric, cartesian coordinates, and integrate both orbits backwards (as test particles) for the total interaction time, $\tint$;
	\item evaluate the likelihood at $t=\tub$.
\end{enumerate}
All model parameters are summarized in Table~\ref{tbl:params}. Assuming each star is an independent tracer, the full posterior for many stars is then just the product of the individual likelihoods with the priors:
\begin{multline}
	p(\Phi, \W, \W_\sat, \tub, \Loffset, \tailbit, \given \D, \D_\sat) = \\
		\shoveleft{\hspace{1em}\frac{1}{\mathcal{Z}} \, \prod_i \, p(\D^{(i)} \given \W^{(i)}) \, p(\D_\sat \given \W_\sat)}\\
			\times p(\W^{(i)} \given \W_\sat, \tub^{(i)}, \Phi, \Loffset, \tailbit^{(i)}) \, 
			\, p(\tub^{(i)}) p(\Phi) \, p(\Loffset)
\end{multline}
This model works well when all stars are lost due to tidal pruning. Even on a mildly eccentric orbit, stars become preferentially unbound around the satellite's pericenter because of tidal shocking (cite?). We will explore properly modeling this process in future work.

@@

We first test the model using eight particles (stars) --- four from the leading tail, and four from the trailing tail --- randomly sampled from the $2.5\times10^8~M_\odot$ simulation (Section~\ref{sec:sims}), assuming perfect observational data for both the stars and the progenitor. The stars were required to have been stripped after the first pericentric passage and have a present-day distance within $50$ kpc of the Sun. Figure~\ref{fig:} (left panel) shows a sample of stream particles (black) and the eight randomly chosen stars (blue) in the Galactic X-Z plane.

We leave the potential parameters --- $q_1$, $q_z$, $\phi$, and $\vhalo$ --- free to vary, along with the Lagrange point offset, $\Loffset$, and use Markov Chain Monte Carlo (MCMC) to sample from the posterior probability distribution over these five parameters. We use an affine-invariant ensemble sampler \citep{goodman10} implemented in the Python programming language \citep{foremanmackey13} with an ensemble of 64 walkers initialized by sampling from the priors summarized in Table~\ref{tbl:params}. We run the walkers in parallel (using MPI todo) for 5000 steps to burn-in, then restart the sampler starting from the final position of the burn-in phase and run for another 10000 steps. We compute the autocorrelation time for each parameter [todo: reference] and thin the chains by taking every XXX$\times \mathrm{median}(t_{\rm acor})$ sample to ensure they are close to independent. Figure~\ref{fig:} shows the walker positions over the 10000 post-burn-in steps for each parameter for the $2.5\times10^8~M_\odot$ simulation; the autocorrelation time for each parameter is displayed on each corresponding panel. Figure~\ref{fig:} shows projections of the posterior probability distribution for the parameters.

@@

For this experiment, we (1) assume the stars are bright distance indicators --- RR Lyrae --- such that the fractional distance uncertainty is $2\%$; (2) neglect the uncertainty in angular position (for a typical RR Lyrae at 50~kpc this is $\sim$$10^{-7}$~deg for \gaia); (3) assume we can measure radial velocities to these stars with 5~km/s uncertainty; and (4) compute the sky-averaged \gaia\, proper-motion uncertainty for each star assuming an F0V spectral type using the \texttt{PyGaia}\footnote{https://github.com/agabrown/PyGaia} code and use this uncertainty for both components of proper motion. The observed positions and velocities of these stars are shown in Figure~\ref{fig:} (red, compare to the unobserved black points). We further assume that we know the tail assignment for each star, $\beta$. We observe the position of the progenitor with the same observational uncertainties, though in reality some coordinates will have higher or lower uncertainties.

This experiment samples over the four potential parameters, the Lagrange point offset, four phase-space coordinate parameters for each star (32 total), and four phase-space coordinate parameters for the progenitor --- 41 parameters in total. We use an ensemble of 256 walkers and draw initial conditions for the coordinate parameters by sampling from Gaussian's centered on the ``observed'' value with variances specified by the observational uncertainties. Other parameters --- potential parameters, $\Loffset$, and $\tub$ --- are initialized by drawing from the priors summarized in Table~\ref{tbl:params}. We again burn in the walkers for 5000 steps and run for 10000 steps post-burn-in. The chains are also thinned by taking every XXX$\times \mathrm{median}(t_{\rm acor})$ sample. Figures~\ref{fig:}-\ref{fig:} show the marginalized posterior for the potential parameters, satellite, and a single particle. [....]

>>>>

% packages
\usepackage{amssymb,amsmath, amsbsy}
\usepackage{booktabs}
\usepackage[caption=false]{subfig}

@@

To model the true, 6D, present-day position of the star, $\W$, we first transform to cartesian, Galactocentric coordinates where ($\bs{r},\bs{v}$) and ($\bs{r}_\sat,\bs{v}_\sat$) are the position and velocity of the star and progenitor. In cartesian coordinates, the present-day position of the star is taken as having been sampled at $t=\tub$ from an isotropic Gaussian centered on one of the instantaneous, effective Lagrange points in position, and an isotropic Gaussian centered on the origin in velocity, and evolved to present day in the parent potential, $\Phi$. We assume that once the star becomes ``unbound'' from the satellite the potential of the satellite can be ignored, and thus treat the star as a test particle.\footnote{This assumption breaks down for sufficiently large-mass progenitors, somewhere between $\sim10^8-10^9~$M$_\odot$.} The satellite mass enters the model through the tidal radius; a changing satellite mass will skew the positions of the effective Lagrange points and velocity scale. We allow for mass-loss by incorporating the initial satellite mass, $m$, and a constant mass-loss term, $\dot{m}$ into the model. Finally, we add two additional parameters: a constant, global offset, $\Loffset$, that scales the position of the effective Lagrange points relative to the classical tidal radius, and a binary parameter $\tailbit$, that is either $-1$ or $+1$ depending on whether the star is in the leading or trailing tail. The likelihood is then:
\begin{align}
	p(\W \given m, \dot{m}, \W_\sat, \tub, \Phi, \Loffset, \tailbit) &= p(\bs{r} \given \bs{r}_\sat, m, \dot{m}, \tub, \Phi, \Loffset, \tailbit)\,p(\bs{v} \given \bs{v}_\sat, m, \dot{m}, \tub, \Phi)\,\left\vert\J\right\vert_{t=\tub}\\
	p(\bs{r} \given \bs{r}_\sat, m, \dot{m}, \tub, \Phi, \Loffset, \tailbit) &= \mathcal{N}(\bs{r} \given \bs{r}_\sat - \Loffset\,\tailbit\,\rtide\,\hat{x}_1, \rtide^2)_{t=\tub}\\
	p(\bs{v} \given \bs{v}_\sat, m, \dot{m}, \tub, \Phi) &= \mathcal{N}(\bs{v} \given \bs{v}_\sat, \sigma^2_v)_{t=\tub}
\end{align}
where $\rtide$ is given by Eq.~\ref{eq:tidalradius}, and $\sigma_v$ by Eq.~\ref{eq:velscale}. $\left\vert\J\right\vert$ is the absolute value of the determinant of the Jacobian that defines the transformation from heliocentric, spherical to Galactocentric, cartesian coordinates. For a single star, in detail, the model works as follows:
\begin{enumerate}
	\item sample from the observed distributions in heliocentric coordinates (Eq.~\ref{eq:obsstar} and Eq.~\ref{eq:obsprog}) to get present-day positions for both the star and progenitor;
	\item transform to Galactocentric, cartesian coordinates, and integrate both orbits backwards (as test particles) for the total interaction time, $\tint$;
	\item evaluate the likelihood at $t=\tub$.
\end{enumerate}
All model parameters are summarized in Table~\ref{tbl:params}. Assuming each star is an independent tracer, the full posterior for many stars is then just the product of the individual likelihoods with the priors:
\begin{multline}
	p(\Phi, \{\W^{(i)}\}, \W_\sat, \tub, \Loffset, \tailbit, \given \{\D^{(i)}\}, \D_\sat) = \\
		\shoveleft{\hspace{1em}\frac{1}{\mathcal{Z}} \, \prod_i \, p(\D^{(i)} \given \W^{(i)}) \, p(\D_\sat \given \W_\sat)}\\
			\times p(\W^{(i)} \given \W_\sat, \tub^{(i)}, \Phi, \Loffset, \tailbit^{(i)}) \, 
			\, p(\tub^{(i)}) p(\Phi) \, p(\Loffset)
\end{multline}
This model works well when all stars are lost due to tidal pruning. Even on a mildly eccentric orbit, stars become preferentially unbound around the satellite's pericenter because of tidal shocking \citep[e.g.,][]{choi09}. We will explore properly modeling this process in future work.

@@

We first test the model using eight particles (stars) --- four from the leading tail, and four from the trailing tail --- randomly sampled from the $2.5\times10^8~M_\odot$ simulation (Section~\ref{sec:sims}), assuming perfect observational data for both the stars and the progenitor. The stars were required to have been stripped after the first pericentric passage and have a present-day distance within $50$ kpc of the Sun. Figure~\ref{fig:2.5e8_xz} (left panel) shows a sample of stream particles (grey) and the eight randomly chosen stars (blue) in the Galactic X-Z plane.

\begin{figure}[h]
\begin{center}
\includegraphics[width=\textwidth]{../../plots/paper2/e8_xz.png}
\caption{ TODO  }\label{fig:2.5e8_xz}
\end{center}
\end{figure}

We leave the potential parameters --- $q_1$, $q_z$, $\phi$, and $\vhalo$ --- free to vary, along with the Lagrange point offset, $\Loffset$, and use Markov Chain Monte Carlo (MCMC) to sample from the posterior probability distribution over these five parameters. We use an affine-invariant ensemble sampler \citep{goodman10} implemented in the Python programming language \citep{foremanmackey13} with an ensemble of 64 walkers initialized by sampling from the priors summarized in Table~\ref{tbl:params}. We run the walkers in parallel (using MPI todo) for 5000 steps to burn-in, then restart the sampler starting from the final position of the burn-in phase and run for another 10000 steps. We compute the autocorrelation time for each parameter [todo: reference] and thin the chains by taking every XXX$\times \mathrm{median}(t_{\rm acor})$ sample to ensure they are close to independent. Figure~\ref{fig:trace} shows the walker positions over the 10000 post-burn-in steps for each parameter for the $2.5\times10^8~M_\odot$ simulation; the autocorrelation time for each parameter is displayed on each corresponding panel. Figure~\ref{fig:exp1_posterior} shows projections of the posterior probability distribution for the parameters.

\begin{figure*}[!ht]
	\captionsetup[subfigure]{labelformat=empty}
        \centering
        \subfloat[]{
                \includegraphics[width=0.5\textwidth]{../../plots/paper2/trace_0.png}
                \label{fig:trace0}
	}
        \subfloat[]{
                \includegraphics[width=0.5\textwidth]{../../plots/paper2/trace_1.png}
                \label{fig:trace0}
	}

        \subfloat[]{
                \includegraphics[width=0.5\textwidth]{../../plots/paper2/trace_2.png}
                \label{fig:trace0}
	}
        \subfloat[]{
                \includegraphics[width=0.5\textwidth]{../../plots/paper2/trace_3.png}
                \label{fig:trace0}
	}

        \subfloat[]{
                \includegraphics[width=0.5\textwidth]{../../plots/paper2/trace_4.png}
                \label{fig:trace0}
	}
        \caption{TODO: mcmc trace plots}\label{fig:trace}
\end{figure*}

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=\textwidth]{../../plots/paper2/exp1_posterior.png}
\caption{ TODO  }\label{fig:exp1_posterior}
\end{center}
\end{figure}

@@

For this experiment, we (1) assume the stars are bright distance indicators --- RR Lyrae --- such that the fractional distance uncertainty is $2\%$; (2) neglect the uncertainty in angular position (for a typical RR Lyrae at 50~kpc this is $\sim$$10^{-7}$~deg for \gaia); (3) assume we can measure radial velocities to these stars with 5~km/s uncertainty; and (4) compute the sky-averaged \gaia\, proper-motion uncertainty for each star assuming an F0V spectral type using the \texttt{PyGaia}\footnote{https://github.com/agabrown/PyGaia} code and use this uncertainty for both components of proper motion. The observed positions and velocities of these stars are shown in Figure~\ref{fig:2.5e8_xz} (red, compare to the unobserved black points). We further assume that we know the tail assignment for each star, $\beta$. We observe the position of the progenitor with the same observational uncertainties, though in reality some coordinates will have higher or lower uncertainties.

This experiment samples over the four potential parameters, the Lagrange point offset, four phase-space coordinate parameters for each star (32 total), and four phase-space coordinate parameters for the progenitor --- 41 parameters in total. We use an ensemble of 256 walkers and draw initial conditions for the coordinate parameters by sampling from Gaussian's centered on the ``observed'' value with variances specified by the observational uncertainties. Other parameters --- potential parameters, $\Loffset$, and $\tub$ --- are initialized by drawing from the priors summarized in Table~\ref{tbl:params}. We again burn in the walkers for 5000 steps and run for 10000 steps post-burn-in. The chains are also thinned by taking every XXX$\times \mathrm{median}(t_{\rm acor})$ sample. Figures~\ref{fig:exp2_potential}, \ref{fig:exp2_satellite}, \ref{fig:exp2_particle0} show the marginalized posteriors for the potential, satellite, and a single particle parameters. [....]

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=\textwidth]{../../plots/paper2/exp2_potential.png}
\caption{ TODO  }\label{fig:exp2_potential}
\end{center}
\end{figure}

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=\textwidth]{../../plots/paper2/exp2_satellite.png}
\caption{ TODO  }\label{fig:exp2_satellite}
\end{center}
\end{figure}

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=\textwidth]{../../plots/paper2/exp2_particle0.png}
\caption{ TODO  }\label{fig:exp2_particle0}
\end{center}
\end{figure}
